### 2010-07-02 doc/guide/language.tex
- made bingo the new gringo trunk
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/trunk/gringo@3149 3e02f596-3131-0410-8352-e04eeaec37cc

The tool \gringo~\cite{gescth07a} is a grounder capable of translating
logic programs provided by users into equivalent ground programs.
The output of \gringo\ can be piped into solver \clasp~\cite{gekanesc07b},
which then computes answer sets.
System \clingo\ internally couples \gringo\ and \clasp, thus,
it takes care of both grounding and solving.
% The input of \gringo\ and \clingo\ is the same,
% but 
In contrast to \gringo\ outputting ground programs, 
\clingo\ returns answer sets.
Both \gringo\ and \clingo\ can handle level-restricted input programs
(cf.\ Section~\ref{subsec:lambda}),
usually specified in one or more text files whose names are passed via the command line
in an invocation of either \gringo\ or \clingo.
We below provide a description of constructs belonging to the input language of \gringo\ and \clingo.

@@

\noindent
The head~$A_0$ of a fact or a rule is an \emph{atom} of the form
$p\code{(}t_1\code{,}\dots\code{,}t_m\code{)}$,
where $p$ is the name of some predicate, that is, a sequence of letters and digits
starting with a lowercase letter,
and any $t_i$ is a term.%
\footnote{An atom~$p$ without arguments is simply a sequence of letters and digits
  (starting with a lowercase letter).
  For such an atom~$p$, parentheses after the name are skipped, e.g.,
  \pred{p42X} could be an atom without arguments.
  Also note that lowercase letters include ``\code{\mus},''
  which can thus be used at any position within a predicate name.}
A \emph{term} $t$ starting with an uppercase letter followed by
a sequence of letters and digits (e.g., \var{X08x15}) is a variable name,
and integers are written as sequences of digits possibly preceded by ``\code{-}.''
In addition, a term can have the same syntactic structure as an atom
(e.g., \code{\pred{p}(\const{a},\const{1},\const{f}(\var{X}))}
 can be either an atom or a term, depending on where it occurs in a rule),
and functions can be nested within a term.
There are also built-in constructs
(cf.\ Section~\ref{subsec:gringo:arith} and~\ref{subsec:gringo:comp})
having particular representations.
Finally, any $L_j$ is a \emph{literal} of the form $A$ or $\code{not}~A$
for an atom~$A$.

@@

@@

Disjunctive logic programs permit connective~``\code{|}'' between atoms in rule heads.
An additional case of translation~$\psi$ from Section~\ref{subsec:semantics}
reflects the semantics of disjunction:
\begin{itemize}
\item $\psi[G\code{|}H]=\psi[G]\vee\psi[H]$.
\end{itemize}
%
The concept of level-restrictedness (cf.\ Section~\ref{subsec:lambda}) is extended
to disjunctive programs by,
for a disjunctive rule~$r$, every predicate~$p/i$
of some atom in the head of~$r$, and each $V\in\vars{r}$,
requiring that there is some $A\in\binders{V}{r}$ such that
$\lambda(p/i)>\lambda(q/j)$ for predicate~$q/j$ of~$A$.
Furthermore, a (maximal) stratified subprogram (cf.\ Section~\ref{subsec:strat})
must not contain any rule with a (non-trivial) disjunctive head.

@@

\gringo\ and \clingo\ support a number of arithmetic functions that
are evaluated during grounding.
The following symbols are used for these functions:
\code{+} (addition),
\code{-} (subtraction),
\code{*} (multiplication),
\code{/} or \code{div} (integer division),
\code{mod} (modulo function),
\code{abs} (absolute value),
\code{\textasciitilde} (bitwise complement),
\code{\&}  (bitwise AND),
\code{?}   (bitwise OR), and
\code{\^} (bitwise exclusive OR).

@@

It is important to note that variables
in the scope of an arithmetic function are not necessarily bound 
(in the sense of Section~\ref{subsec:lambda}) by a corresponding atom.
For instance, atom \code{\pred{p}(\var{X}+1,\var{X})} belongs to
$\binders{\var{X}}{r}$, while atom \code{\pred{p}(\var{X}+1,\var{Y})}
does not belong to $\binders{\var{X}}{r}$ because it contains~\var{X}
only in the scope of an arithmetic function.

@@

Importantly, a built-in comparison predicate cannot bind variables,
that is, it does not belong to $\binders{\var{X}}{r}$ for any variable~\var{X}.
Also note that built-in predicates are not considered by level mappings~$\xi$
in the context of stratification (cf.\ Section~\ref{subsec:strat}), or equivalently,
one may assume that built-in predicates are mapped to level~$0$ and all other
predicates to levels greater than~$0$.

@@

Built-in predicate~\code{=} can be used in the body of a rule
(or on right-hand sides of conditions introduced in Section~\ref{subsec:gringo:condition})
to assign a term on its right-hand side to a variable on its left-hand side.
Note that ``\code{\var{X} \!=\! }$t$'' belongs to~$\binders{\var{X}}{r}$,
provided that~$t$ is variable-free or that its variables are bound
by atoms other than assignments with~\var{X} on the right-hand side.
Cyclic assignments over otherwise unbound variables are thus excluded.

\begin{example}\label{ex:assign}
The next program demonstrates how terms can be assigned to variables:%
\marginlabel{%
  The unique answer set of the program is obtained via call:\\
  \code{\mbox{~}gringo -t \textbackslash\\\mbox{~}examples/assign.lp}}
%
\lstinputlisting{examples/assign.lp}
%
Line~3 contains four assignments, where the right-hand sides directly or indirectly
depend on~\var{X} and~\var{Y}.
These two variables are bound in Line~4 via atoms of predicate \pred{num}/$1$.
Also observe the different usage and role of built-in comparison predicate~\code{==}.
\eexample
\end{example}

@@

As with built-in arithmetic functions,
an integer interval mentioning some variable (like~\var{X} in Line~4 of Example~\ref{ex:int})
cannot be used to bind the variable.

@@

Conditions allow for instantiating variables to
collections of terms within a single rule.
This is particularly useful for encoding conjunctions or disjunctions over
arbitrarily many ground atoms as well as for the compact representation of aggregates
(cf.\ Section~\ref{subsec:gringo:aggregate}).
The symbol~``\code{:}'' is used to formulate conditions.

@@

Symbol~``\code{;}'' allows for pooling alternative terms to be used as
argument within an atom, thus, specifying rules more compactly.
An atom written in the form \code{\pred{p}($\dots$,X;Y,$\dots$)}
abbreviates two options:
\code{\pred{p}($\dots$,X,$\dots$)} and \code{\pred{p}($\dots$,Y,$\dots$)}.
Pooled arguments in an atom of a rule body 
(or on the right-hand side of a condition) are expanded to a conjunction
of the options within the same body (or within the same condition),
while they are expanded to multiple rules 
(or multiple literals connected via ``\code{,}'')
when occurring in the head (or in front of a condition).

\begin{example}\label{ex:pool}
The following logic program makes use of pooling:
%
\lstinputlisting{examples/pool.lp}
%
Let us consider instantiations of the rule in Line~3 obtained with substitution 
$\{\var{A}\mapsto\const{a},\linebreak[1]\var{B}\mapsto\const{b},
   \var{M}\mapsto\const{1},\var{N}\mapsto\const{2}\}$.
Note that \pred{mix}/$2$ and \pred{-mix}/$2$ each admit four options,
corresponding to the cross product of $\{\const{a},\const{b}\}$ substituted
for~\var{A} and~\var{B}, respectively, together with $\{\const{1},\const{2}\}$
substituted for~\var{M} and~\var{N}.
While the instances obtained for \pred{mix}/$2$ give rise to four rules,
the instances for \pred{-mix}/$2$ jointly belong to the body.
The (repeated) body also contains two instances each of \pred{sym}/$1$ and of \pred{num}/$1$.
We thus get the rules:%
\marginlabel{%
  Simplified versions of these rules are produced via call:\\
  \code{\mbox{~}gringo -t \textbackslash\\\mbox{~}examples/pool.lp}}
%
\begin{lstlisting}[numbers=none]
mix(a,1) :- sym(a),sym(b), num(1),num(2), not -mix(1,a),
  not -mix(1,b), not -mix(2,a), not -mix(2,b).
mix(a,2) :- sym(a),sym(b), num(1),num(2), not -mix(1,a),
  not -mix(1,b), not -mix(2,a), not -mix(2,b).
mix(b,1) :- sym(a),sym(b), num(1),num(2), not -mix(1,a),
  not -mix(1,b), not -mix(2,a), not -mix(2,b).
mix(b,2) :- sym(a),sym(b), num(1),num(2), not -mix(1,a),
  not -mix(1,b), not -mix(2,a), not -mix(2,b).
\end{lstlisting}
%
Finally, we note that pooling is also possible with arguments of built-in predicates.
\eexample
\end{example}

\subsubsection{Aggregates}\label{subsec:gringo:aggregate}

An aggregate is an operation on a multiset of weighted literals that evaluates to an integer.
In combination with arithmetic comparisons, we can extract a truth value from
an aggregate's evaluation, thus, obtaining an aggregate atom.
Customizing the notation in~\cite{ferraris05a},
we consider ground \emph{aggregate atoms} of the form:
%
\begin{equation}\label{eq:aggregate:atom}
  l~ 
  \mathrm{op} 
  \code{[}
  L_1\code{=}w_1\code{,}
  \dots\code{,}
  L_n\code{=}w_n
  \code{]}
  u
\end{equation}
%
where~$l$ and~$u$ are integers each of which can possibly be omitted,
$\mathrm{op}$ is a function from multisets of integers to an integer,
each $L_i$ is a literal of the form~$A$ or $\code{not}~A$
for some atom~$A\in\ground{\mathcal{A}}$,
and each~$w_i$ is an integer.
The intuitive reading of an aggregate atom is that the value returned by~$\mathrm{op}$
applied to all weights~$w_i$ such that~$L_i$ holds should be in-between
lower bound~$l$ and upper bound~$u$ (inclusively).
If~$l$ or~$u$ is omitted,
it means that the aggregate's value is not constrained from below or above,
respectively.

For an aggregate atom~$\mathit{Agg}$ as in~(\ref{eq:aggregate:atom}),
we let $\atom{\mathit{Agg}}=\{L_i\in\nolinebreak\ground{\mathcal{A}}\mid 1\leq i\leq n\}$
denote the set of atoms occurring positively in~$\mathit{Agg}$.
Currently, \gringo\ and \clingo\ support aggregate operations
\const{sum}, \const{max}, and \const{min}.%
\footnote{%
  The additional operation \const{count} is a shorthand for a \const{sum}
  aggregate in which all weights are~\const{1}.}
We can now extend translations~$\phi$ and~$\psi$ from Section~\ref{subsec:semantics}
to aggregate atoms, 
using the approach in~\cite{ferraris05a} but
applying the ``choice semantics'' for aggregates
in heads of rules~\cite{ferlif05a,siniso02a}:%
\footnote{%
  The translation~$\phi$ we provide for \const{sum} aggregates
  assumes that there are no negative weights for literals,
  while the more sophisticated translation in~\cite{ferraris05a} also
  captures negative weights.
  All current solvers work with positive weights only, and thus
  \gringo\ and \lparse\ use a compilation technique~\cite{siniso02a}
  to eliminate negative weights, which are not permitted in
  \lparse's output format~\cite{lparseManual}.
  As this technique may lead to counterintuitive results~\cite{ferraris05a},
  we strongly recommend not to use negative weights within \const{sum} aggregates.}
\begin{enumerate}
\item % op %
$\phi[
  \mathrm{op} 
  \code{[}
  L_1\code{=}w_1\code{,}
  \dots\code{,}
  L_n\code{=}w_n
  \code{]}
  ] = 
 \top$,
\item % l op u %
$\begin{array}[t]{@{}r@{}l@{}}
  \phi[
  l~
 &
  \mathrm{op} 
  \code{[}
  L_1\code{=}w_1\code{,}
  \dots\code{,}
  L_n\code{=}w_n
  \code{]}
  u
  ] = 
 {} \\ 
 &
  \phi[
  l~
  \mathrm{op} 
  \code{[}
  L_1\code{=}w_1\code{,}
  \dots\code{,}
  L_n\code{=}w_n
  \code{]}
  ]
  \wedge
  \phi[
  \mathrm{op} 
  \code{[}
  L_1\code{=}w_1\code{,}
  \dots\code{,}
  L_n\code{=}w_n
  \code{]}
  u
  ]\text{,}
 \end{array}$
\item % l sum %
$\begin{array}[t]{@{}r@{}l@{}}
  \phi[
  l~
 &
  \const{sum} 
  \code{[}
  L_1\code{=}w_1\code{,}
  \dots\code{,}
  L_n\code{=}w_n
  \code{]}
  ] = 
 {} \\ 
 &
  \big(
  \bigvee_{\{i_1,\dots,i_j\}\subseteq\{1,\dots,n\},l\leq w_{i_1}+\dots+w_{i_j}}
  (\phi[L_{i_1}]\wedge\dots\wedge \phi[L_{i_j}])
  \big)\text{,}
 \end{array}$
\item % sum u %
$\begin{array}[t]{@{}r@{}l@{}}
  \phi
  [
 &
  \const{sum} 
  \code{[}
  L_1\code{=}w_1\code{,}
  \dots\code{,}
  L_n\code{=}w_n
  \code{]}
  u
  ] = 
 {} \\ 
 &
  \neg
  \big(
  \bigvee_{\{i_1,\dots,i_j\}\subseteq\{1,\dots,n\},u< w_{i_1}+\dots+w_{i_j}}
  (\phi[L_{i_1}]\wedge\dots\wedge \phi[L_{i_j}])
  \big)\text{,}
 \end{array}$
\item % l max %
 $\phi[
  l~
  \const{max} 
  \code{[}
  L_1\code{=}w_1\code{,}
  \dots\code{,}
  L_n\code{=}w_n
  \code{]}
  ] = 
 \hspace{6pt}
  \big(
  \bigvee_{i\in\{1,\dots,n\},l\leq w_i}
  \phi[L_i]
  \big)$,
\item % max u %
 $\phi[
  \const{max} 
  \code{[}
  L_1\code{=}w_1\code{,}
  \dots\code{,}
  L_n\code{=}w_n
  \code{]}
  u
  ] = 
  \neg
  \big(
  \bigvee_{i\in\{1,\dots,n\},u< w_i}
  \phi[L_i]
  \big)$,
\item % l min %
 $\phi[
  l~
  \const{min} 
  \code{[}
  L_1\code{=}w_1\code{,}
  \dots\code{,}
  L_n\code{=}w_n
  \code{]}
  ] = 
  \neg
  \big(
  \bigvee_{i\in\{1,\dots,n\},w_i< l}
  \phi[L_i]
  \big)
 $,
\item % min u %
 $\phi[
  \const{min} 
  \code{[}
  L_1\code{=}w_1\code{,}
  \dots\code{,}
  L_n\code{=}w_n
  \code{]}
  u
  ] = 
  \hspace{6pt}
  \big(
  \bigvee_{i\in\{1,\dots,n\},w_i\leq u}
  \phi[L_i]
  \big)
 $, and
\item % psi %
$\begin{array}[t]{@{}r@{}l@{}}
  \psi[
  \lhd
 &~
  \mathrm{op} 
  \code{[}
  L_1\code{=}w_1\code{,}
  \dots\code{,}
  L_n\code{=}w_n
  \code{]}
  \rhd
  ] = 
% {} \\ 
% &
   \phi[
  \lhd~
  \mathrm{op} 
  \code{[}
  L_1\code{=}w_1\code{,}
  \dots\code{,}
  L_n\code{=}w_n
  \code{]}
  \rhd
  ]
  \wedge
 {} \\ 
 &
  \big(
  \bigwedge_{A\in\atom{
    \lhd~
    \mathrm{op} 
    \code{[}
    L_1\code{=}w_1\code{,}
    \dots\code{,}
    L_n\code{=}w_n
    \code{]}
    \rhd
  }}
  (A\vee\neg A)
  \big)\text{.}
 \end{array}$
\end{enumerate}
%
The first item reflects that an unconstrained aggregate atom always holds,
while aggregate atoms with a lower and an upper bound are split up into two
conjunctively connected parts (Item~2).
Item~3--8 specify these parts individually for~\const{sum},~\const{max}, and~\const{min}
aggregates, respectively.
Provided that only non-negative weights are used within \const{sum}s,
all three aggregates are subject to one comparison that is monotone,
meaning that by default it evaluates to \textit{false} so that truth must be established.
For \const{sum} and \const{max} aggregates, these monotone comparisons are related
to lower bounds (Item~3 and~5),
while an upper bound behaves monotonically for \const{min} (Item~8).
In turn, the opposite comparison is antimonotone, that is,
by default it evaluates to \textit{true} unless too many or improper literals hold.
This is the case for comparisons to upper bounds with aggregates
\const{sum} and \const{max} (Item~4 and~6), while it concerns 
the lower bound for \const{min} (Item~7).
Finally, translation~$\psi$ of an aggregate atom in the head of a rule consists of two parts:
first, $\phi$ is used like in the body of a rule
(in Item~9, $\lhd$ and~$\rhd$ are used as ``wildcards'' for provided or omitted 
 lower and upper bounds),
second, a disjunction $(A\vee\neg A)$ is added for each atom~$A$ occurring positively.
Even though such disjunctions are tautological,
they are important under answer set semantics (cf.\ Section~\ref{subsec:semantics}),
as they permit deriving an arbitrary subset of atoms from an aggregate in the head
(which explains the name  ``choice semantics'').
Beyond this special treatment in rule heads,
the truth values of aggregate atoms result from their literals as
may be intuitively expected,
and Item~1--8 basically do nothing but formalizing the standard meanings
of aggregate operations.
Note that the above translations merely define the semantics of aggregates,
which does not imply that solvers unfold them in this way.
In fact, \clasp\ natively supports aggregates without translating them.

As regards syntactic representation,
weight~\const{1} is considered a default,
so that $L_i\code{=}\const{1}$ can simply be written as $L_i$.
For instance, the following (multi)sets of (weighted) literals are the same
when combined with any kind of aggregate operation and bounds:
\begin{lstlisting}[numbers=none,escapechar=@]
[a=1, not b=1, c=2]  @\textnormal{and}@
[a,   not b,   c=2]@\textnormal{.}@
\end{lstlisting}
%
Furthermore,
keyword \const{sum} may be omitted, which in a sense makes \const{sum} the default
aggregate operation.
In fact, the following aggregate atoms are synonyms:
\begin{lstlisting}[numbers=none,escapechar=@]
2 sum [a, not b, c=2] 3  @\textnormal{and}@
2     [a, not b, c=2] 3@\textnormal{.}@
\end{lstlisting}
By omitting keyword \const{sum}, we obtain the same notation as the one of 
so-called ``weight constraints''~\cite{siniso02a,lparseManual}, which are actually aggregate atoms
whose operation is addition.

It is important to note that the (weighted) literals within an aggregate
belong to a multiset.
In particular, if there are multiple occurrences
$L\code{=}w_1,\dots,L\code{=}w_k$
of a literal~$L$, in combination with \const{min} and \const{max},
it is not the same like having $L\code{=}w_1+\dots+w_k$.
To see this, note that the program consisting of the facts:
\begin{lstlisting}[numbers=none]
2 max [a=2].  2 min [a=2].
\end{lstlisting}
has $\{\code{a}\}$ as its unique answer set, while there is no answer set for:
\begin{lstlisting}[numbers=none]
2 max [a,a].  2 min [a,a].
\end{lstlisting}

If literals ought not to be repeated,
we can use \const{count} instead of \const{sum}.
Syntactically, \const{count} requires curly instead of square
brackets, and there must not be any weights within a \const{count} aggregate.
Regarding semantics,
% \begin{equation*}%\label{eq:aggregate:atom}
$(
  l~ 
  \const{count} 
  \code{\char`\{}
  L_1\code{,}
  \dots\code{,}
  L_n
  \code{\char`\}}
  u
)$ 
%\end{equation*}
reduces to
$(
  l~ 
  \const{sum} 
  \code{[}
  L_1\code{=}\const{1}\code{,}
  \dots\code{,}
  L_m\code{=}\const{1}
  \code{]}
  u
)$,
where $\{L_1,\dots,L_m\}=\{L_i \mid 1\leq i\leq n\}$ is obtained by dropping
repeated literals.
Of course, the use of~$l$ and~$u$ is optional also with \const{count}.
As an example, note that the next aggregate atoms express the same:
\begin{lstlisting}[numbers=none,escapechar=@]
1 sum   [a=1, not b=1]     1  @\textnormal{and}@ 
1 count {a,a, not b,not b} 1@\textnormal{.}@  
\end{lstlisting}
Keyword \const{count} can be omitted (like \const{sum}),
so that the following are synonyms:
\begin{lstlisting}[numbers=none,escapechar=@]
1 count {a, not b} 1  @\textnormal{and}@ 
1       {a, not b} 1@\textnormal{.}@  
\end{lstlisting}
The last notation is similar to the one of so-called
``cardinality constraints''~\cite{siniso02a,lparseManual},
which are aggregate atoms using counting as their operation.

After considering the syntax and semantics of ground aggregate atoms,
we now turn our attention to non-ground aggregates.
Regarding contained variables, an atom occurring in an aggregate behaves
similar to an atom on the left-hand side of a condition
(cf.\ Section~\ref{subsec:gringo:condition}).
That is, any variable occurring within an aggregate is a priori local,
and it must be bound via a variable of the same name that is global
or that occurs on the right-hand side of a condition
(with the atom containing the variable in front).
As with local variables of conditions, global variables take priority
during grounding, so that the names of local variables must be chosen
with care to avoid accidental clashes.
Beyond conditions (which are more or less the natural construct to use
for instantiating variables within an aggregate),
classical negation (cf.\ Section~\ref{subsec:gringo:negation}),
built-in arithmetic functions (cf.\ Section~\ref{subsec:gringo:arith}),
intervals (cf.\ Section~\ref{subsec:gringo:interval}), and 
pooling (cf.\ Section~\ref{subsec:gringo:pool})
can be incorporated as usual within aggregates,
where intervals and pooling are expanded locally.%
\footnote{%
  Assignments (cf.\ Section~\ref{subsec:gringo:assign}) and (currently)
  also built-in comparison predicates (cf.\ Section~\ref{subsec:gringo:comp})
  are permitted on the right-hand side of conditions only.}
That is, an interval gives rise to multiple literals connected via~``\code{,}''
within the same aggregate.
The same applies to pooling in front of a condition,
while it turns into a composite condition chained by~``\code{:}'' on the right-hand side.
The notions of level-restrictedness and stratification
(cf.\ Section~\ref{subsec:lambda} and~\ref{subsec:strat}) are extended
to programs with aggregates by
considering the predicates of all atoms (except for those on the right-hand side of a condition)
in an aggregate atom occurring as the head of a rule~$r$,%
\footnote{%
  The rules in a (maximal) stratified subprogram (cf.\ Section~\ref{subsec:strat})
  cannot have aggregate atoms in the head.}
and by assuming $A\in\nbody{r}$ for all atoms~$A$ of an aggregate in the body of~$r$.
Finally, note that aggregates without bounds are also permitted on the
right-hand sides of assignments, but using this feature is only recommended
for aggregates whose atoms belong to domain predicates because space blow-up
can become a bottleneck otherwise.
The following example, making exhaustive use of aggregates,
nonetheless demonstrates this and other features.

\begin{example}\label{ex:aggr}
Consider a situation where an informatics student
wants to enroll for a number of courses at the beginning of a new term.
In the university calendar, eight courses are found eligible,
and they are represented by the following facts:
%
\lstinputlisting[xrightmargin=-20pt,lastline=8]{examples/aggr.lp}
%
In an instance of \const{course}/$3$,
the first argument is a number identifying one of the eight courses,
and the third argument provides the course's contact hours per week.
The second argument stands for a subject area:
\const{1} corresponding to ``theoretical informatics,''
\const{2}               to ``practical informatics,''
\const{3}               to ``technical informatics,''
and~\const{4}               to ``applied informatics.''
For instance, atom \code{\const{course}(\const{1},\const{2},\const{5})}
expresses that course~\const{1} accounts for~\const{5} contact hours per week
that may be credited to subject area~\const{2} (``practical informatics'').
Observe that a single course is usually eligible for multiple
subject areas.

After specifying the above facts,
the student starts to provide personal constraints on the courses to enroll.
The first condition is that~\const{3} to~\const{6} courses should be enrolled:
%
\lstinputlisting[nolol,firstline=11,lastline=11,firstnumber=9]{examples/aggr.lp}
%
Instantiating the above \const{count} aggregate
yields the following ground rule:%
\marginlabel{%
  The full ground program is obtained by invoking:\\
  \code{\mbox{~}gringo -t \textbackslash\\\mbox{~}examples/aggr.lp}}
%
\begin{lstlisting}[firstnumber=9,stepnumber=9]
3 { enroll(1), enroll(2), enroll(3), enroll(4), 
    enroll(5), enroll(6), enroll(7), enroll(8) } 6.
\end{lstlisting}
%
Observe that an instance of atom~\code{\const{enroll}(\var{C})} is included for each
instantiation of~\var{C} such that \code{\const{course}(\var{C},\var{S},\var{H})}
holds for some values of~\var{S} and~\var{H}.
Duplicates resulting from distinct values for~\var{S} are removed, thus,
obtaining the above set of ground atoms.

>>>>

The tool \gringo~\cite{gescth07a} is a grounder capable of translating
logic programs provided by users into equivalent ground programs.
The output of \gringo\ can be piped into solver \clasp~\cite{gekanesc07b},
which then computes answer sets.
System \clingo\ internally couples \gringo\ and \clasp, thus,
it takes care of both grounding and solving.
% The input of \gringo\ and \clingo\ is the same,
% but 
In contrast to \gringo\ outputting ground programs, 
\clingo\ returns answer sets.
Both \gringo\ and \clingo\ can handle safe input programs
(cf.\ Section~\ref{subsec:lambda}),
usually specified in one or more text files whose names are passed via the command line
in an invocation of either \gringo\ or \clingo.
We below provide a description of constructs belonging to the input language of \gringo\ and \clingo.

@@

\noindent
The head~$A_0$ of a fact or a rule is an \emph{atom} of the form
$p\code{(}t_1\code{,}\dots\code{,}t_m\code{)}$,
where $p$ is the name of some predicate, that is, a sequence of letters, digits and underscores 
starting with a lowercase letter,
and any $t_i$ is a term.%
\footnote{An atom~$p$ without arguments is simply a sequence of letters and digits
  (starting with a lowercase letter).
  For such an atom~$p$, parentheses after the name are skipped, e.g.,
  \pred{p42X} could be an atom without arguments.}
A \emph{term} $t$ starting with an uppercase letter followed by
a sequence of letters, digits and underscores (e.g., \var{X\_08x15}) is a variable name,
and integers are written as sequences of digits possibly preceded by ``\code{-}.''
In addition, a term can have the same syntactic structure as an atom
(e.g., \code{\pred{p}(\const{a},\const{1},\const{f}(\var{X}))}
 can be either an atom or a term, depending on where it occurs in a rule),
and functions can be nested within a term.
There are also built-in constructs
(cf.\ Section~\ref{subsec:gringo:arith} and~\ref{subsec:gringo:comp})
having particular representations.
Finally, any $L_j$ is a \emph{literal} of the form $A$ or $\code{not}~A$
for an atom~$A$.

@@

\begin{newstuff}
\subsubsection{Anonymous Variables}
Variables that start with an underscore are so called \emph{anonymous variables}.
Each such variable is treated like a distinct variable name.
They are especially helpful to mark places where we do not care about the value of the variable.
\begin{example}
The following example demonstrates that such variables come in handy
if we want to project out some variables.
\begin{lstlisting}
p(1,1,1).
p(2,1,2).
p(X) :- p(X,_A,_A).
\end{lstlisting}
Note that both occurrences of \code{\_A} are treated as distinct variables (we get both~\code{p(1)} and~\code{p(2)}).
\end{example}
\end{newstuff}

@@

Disjunctive logic programs permit connective~``\code{|}'' between atoms in rule heads.
An additional case of translation~$\psi$ from Section~\ref{subsec:semantics}
reflects the semantics of disjunction:
\begin{itemize}
\item $\psi[G\code{|}H]=\psi[G]\vee\psi[H]$.
\end{itemize}
%
The concept of safe programs (cf.\ Section~\ref{subsec:safe}) directly applies
to disjunctive programs.
Furthermore, a (maximal) stratified subprogram (cf.\ Section~\ref{subsec:strat})
must not contain any rule with a (non-trivial) disjunctive head.

@@

\gringo\ and \clingo\ support a number of arithmetic functions that
are evaluated during grounding.
The following symbols are used for these functions:
\code{+} (addition),
\code{-} (subtraction),
\code{*} (multiplication),
\code{/} or \code{\#div} (integer division),
\code{\#mod} (modulo function),
\code{\#abs} (absolute value), and
\code{\#pow} (power function).

@@

\begin{newstuff}
It is important to note that variables
in the scope of an arithmetic function are not bound 
(in the sense of Section~\ref{subsec:safe}) by a corresponding (positive) atom.
Only edges for variables that are not in the scope of arithmetic functions
are added to the variable graph~$G=(V,E)$.
For instance, atom \code{\pred{p}(\var{X}+1,\var{X})} provides edge~$(\$, X)$,
while atom \code{\pred{p}(\var{X}+1,\var{Y})} provides edge~$(\$, Y)$,
but does not provide edge~$(\$,X)$ because it contains~\var{X}
only in the scope of an arithmetic function.

Note that previous version of \gringo\ also supported the arithmetic functions
\code{\textasciitilde} (bitwise complement),
\code{\&}  (bitwise AND),
\code{?}   (bitwise OR), and
\code{\^} (bitwise exclusive OR).
These have been removed because there are scarcely application applications using them.
Furthermore, with \gringo's new Lua support\comment{Reference to lua section}\ arbitrary arithmetic functions can be implemented.
\end{newstuff}

@@

\begin{newstuff}
Importantly, built-in comparison predicates do not provide edges to variable graphs (see Section~\ref{subsec:safe}).
\end{newstuff}

@@

Built-in predicate~\code{:=}%
\footnote{Note that there is also the built-in predicate~\code{=},
which, in contrast to~\code{:=}, only permits variables on its left hand side.
Apart from that both literals are semantically the same.
Predicate~\code{=} is mainly kept for compatibility with \lparse{} and older \gringo{} versions.} 
can be used in the body of a rule
(or on right-hand sides of conditions introduced in Section~\ref{subsec:gringo:condition})
to unify a term on its right-hand side to a term on its left-hand side.

%Note that the variable~$X$ in ``\code{\var{X} \!=\! }$t$'' is treated like a positive occurence if
%the variables on the right-hand side are bound positively minus cyclic assignments.
\begin{example}\label{ex:assign}
The next program demonstrates how terms can be assigned to variables:%
\marginlabel{%
  The unique answer set of the program is obtained via call:\\
  \code{\mbox{~}gringo -t \textbackslash\\\mbox{~}examples/assign.lp}}
%
\lstinputlisting{examples/assign.lp}
%
Line~3 contains four assignments, where the right-hand sides directly or indirectly
depend on~\var{X} and~\var{Y}.
These two variables are bound in Line~4 via atoms of predicate \pred{num}/$1$.
Also observe the different usage and role of built-in comparison predicate~\code{==}.
\eexample
\end{example}

\begin{newstuff}
For an assignemt literal~$A$, the variable graph (see Section~\ref{subsec:safe}) contains edges from 
the variables of the right hand sight to the variables of the left hand side of~$A$.
If a variable on the right hand side occurs in the scope of an arithmetic function,
then no edge is added.
If the right hand side does not contain variables then edges start with~$\$$.
For instance, literal \code{f(X,X+Y):=f(U,V)} provides the edges $(U,X)$ and $(V,X)$.
Intuitively, this means the assignment can be used to bind $X$ if $U$ and $V$ are bound by some other literal.
\end{newstuff}

@@

\begin{newstuff}
As with built-in arithmetic functions,
an integer interval mentioning some variable (like~\var{X} in Line~4 of Example~\ref{ex:int})
does not contribute edges in the variable graph, i.e., it will never bind variables.
\end{newstuff}

@@

Conditions allow for instantiating variables to
collections of terms within a single rule.
This is particularly useful for encoding conjunctions%
\footnote{Note that conditions also permit comparison predicates in front of the first colon.}
or disjunctions over
arbitrarily many ground atoms as well as for the compact representation of aggregates
(cf.\ Section~\ref{subsec:gringo:aggregate}).
The symbol~``\code{:}'' is used to formulate conditions.

@@

Symbol~``\code{;}'' allows for pooling alternative terms to be used as
argument within an atom, thus, specifying rules more compactly.
An atom written in the form \code{\pred{p}($\dots$,X;Y,$\dots$)}
abbreviates two options:
\code{\pred{p}($\dots$,X,$\dots$)} and \code{\pred{p}($\dots$,Y,$\dots$)}.
Pooled arguments in an atom of a rule body 
(or on the right-hand side of a condition) are expanded to a conjunction
of the options within the same body (or within the same condition),
while they are expanded to multiple rules 
(or multiple literals connected via~``\code{,}'')
when occurring in the head (or in front of a condition).

\begin{example}\label{ex:pool}
The following logic program makes use of pooling:
%
\lstinputlisting{examples/pool.lp}
%
Let us consider instantiations of the rule in Line~3 obtained with substitution 
$\{\var{A}\mapsto\const{a},\linebreak[1]\var{B}\mapsto\const{b},
   \var{M}\mapsto\const{1},\var{N}\mapsto\const{2}\}$.
Note that \pred{mix}/$2$ and \pred{-mix}/$2$ each admit four options,
corresponding to the cross product of $\{\const{a},\const{b}\}$ substituted
for~\var{A} and~\var{B}, respectively, together with $\{\const{1},\const{2}\}$
substituted for~\var{M} and~\var{N}.
While the instances obtained for \pred{mix}/$2$ give rise to four rules,
the instances for \pred{-mix}/$2$ jointly belong to the body.
The (repeated) body also contains two instances each of \pred{sym}/$1$ and of \pred{num}/$1$.
We thus get the rules:%
\marginlabel{%
  Simplified versions of these rules are produced via call:\\
  \code{\mbox{~}gringo -t \textbackslash\\\mbox{~}examples/pool.lp}}
%
\begin{lstlisting}[numbers=none]
mix(a,1) :- sym(a),sym(b), num(1),num(2), not -mix(1,a),
  not -mix(1,b), not -mix(2,a), not -mix(2,b).
mix(a,2) :- sym(a),sym(b), num(1),num(2), not -mix(1,a),
  not -mix(1,b), not -mix(2,a), not -mix(2,b).
mix(b,1) :- sym(a),sym(b), num(1),num(2), not -mix(1,a),
  not -mix(1,b), not -mix(2,a), not -mix(2,b).
mix(b,2) :- sym(a),sym(b), num(1),num(2), not -mix(1,a),
  not -mix(1,b), not -mix(2,a), not -mix(2,b).
\end{lstlisting}
%
Finally, we note that pooling is possible in every term.
Although, it is most useful in predicates.
\eexample
\end{example}

\begin{newstuff}
There is a second pooling operator \code{;;} that is only permitted in predicates.
It is used to seperate arguments of predicates,
e.g., \code{p(a,b;;c,d)} expands to \code{p(a,b), p(c,d)}.
Note that this operator corresponds to \lparse's~\code{;} operator.
\end{newstuff}

\subsubsection{Aggregates}\label{subsec:gringo:aggregate}
\input{aggregates}
